---
title: "Getting Started with rsparrow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rsparrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**rsparrow** is an R package for executing and evaluating SPARROW (SPAtially Referenced
Regressions On Watershed attributes) models. SPARROW models three watershed processes:

1. **Source generation** — contaminant inputs from land uses (e.g., agricultural N,
   atmospheric deposition)
2. **Land-to-water delivery** — transport to streams via surficial and subsurface pathways,
   modulated by climate and landscape variables
3. **In-stream transport and decay** — removal and attenuation as water moves downstream
   through reaches and reservoirs

The package implements static (long-term mean annual) SPARROW models. It generates
reach-level load and yield predictions, source attribution, parameter uncertainty via
bootstrap, and diagnostic plots.

## Installation

```{r install, eval=FALSE}
install.packages("rsparrow")

# Or the development version from GitHub:
# remotes::install_github("Kyle-Hurley/rsparrow")
```

## Step 1: Load the Example Data

rsparrow includes a synthetic 60-reach example dataset, `sparrow_example`, that
demonstrates the full modeling workflow without requiring real data.

```{r load-data}
library(rsparrow)

str(sparrow_example, max.level = 1)
```

The dataset contains five elements:

- **`reaches`** — 60-reach dendritic network with two tributary branches draining to
  a single outlet. Columns include topology (`waterid`, `fnode`, `tnode`), drainage
  area (`demiarea`, `demtarea`), a synthetic agricultural-N source (`agN`), and a
  precipitation delivery variable (`ppt`).
- **`sites`** — 5 monitoring sites with synthetic observed loads (`depvar`, kg/year).
- **`parameters`** — parameter configuration (SOURCE and DELIVF types).
- **`design_matrix`** — binary matrix associating sources with delivery variables.
- **`data_dictionary`** — maps internal SPARROW variable names to data column names.

## Step 2: Inspect the Network Topology

Before fitting a model, compute the hydrological sequence — the reach ordering from
upstream to downstream that SPARROW uses for load accumulation.

```{r hydseq}
network <- rsparrow_hydseq(sparrow_example$reaches)
head(network[order(network$hydseq), c("waterid", "fnode", "tnode", "hydseq")])
```

Reaches with more negative `hydseq` values are further downstream. The outlet reach
has the most negative value.

## Step 3: Fit the Model

Pass the four data frames directly to `rsparrow_model()`. No CSV files or directories
are required:

```{r fit-model, eval=FALSE}
model <- rsparrow_model(
  sparrow_example$reaches,
  sparrow_example$parameters,
  sparrow_example$design_matrix,
  sparrow_example$data_dictionary
)
```

After a few seconds, the model produces a summary of estimation performance:

```
  SPARROW NLLS MODEL SUMMARY

  MODEL NAME: run1

  MODEL ESTIMATION PERFORMANCE (Monitoring-Adjusted Predictions)
 MOBS NPARM DF       SSE       MSE      RMSE       RSQ
    5     2  3 0.3410997 0.1136999 0.3371942 0.8056507

  PARAMETER ESTIMATES
  PARAMETER PARM TYPE ESTIMATE     SE         T         P-VALUE
1 agN       SOURCE    3.417676  0.6478    5.276    0.01327
2 ppt       DELIVF   -0.001593  0.0038   -0.422    0.70122
```

## Step 4: Examine Results

The returned `rsparrow` object supports standard S3 methods:

```{r examine, eval=FALSE}
# Compact summary
print(model)

# Full estimation summary
summary(model)

# Parameter estimates
coef(model)
#>        agN        ppt
#>  3.4176761 -0.0015935

# Log residuals at calibration sites
residuals(model)

# Parameter covariance matrix
vcov(model)

# Predictions (populated when if_predict = TRUE)
names(model$predictions)
#> [1] "oparmlist"          "loadunits"          "predmatrix"
#> [4] "oyieldlist"         "yieldunits"         "yldmatrix"
```

## Step 5: Diagnostic Plots

If the **plotly** package is installed, diagnostic residual plots are available:

```{r plots, eval=FALSE}
plot(model, type = "residuals")           # observed vs predicted, residuals
plot(model, type = "residuals", panel = "B")  # Q-Q, boxplots
plot(model, type = "sensitivity")         # parameter sensitivity
plot(model, type = "spatial")             # spatial autocorrelation
```

## Step 6: Bootstrap Uncertainty

Estimate parameter uncertainty via parametric bootstrap:

```{r bootstrap, eval=FALSE}
model <- rsparrow_bootstrap(model, n_boot = 200, seed = 42)
model$bootstrap$bEstimate  # bootstrapped parameter distributions
```

## Step 7: Scenario Analysis

Evaluate the effect of changing source inputs:

```{r scenario, eval=FALSE}
# Simulate a 50% reduction in agricultural N loading
model <- rsparrow_scenario(model, source_changes = list(agN = 0.5))
```

## Further Reading

- `?rsparrow_model` — full parameter documentation
- `?sparrow_example` — description of the example dataset
- SPARROW methodology: Alexander, R.B., et al. (2008). *Differences in phosphorus and
  nitrogen delivery to the Gulf of Mexico from the Mississippi River basin.*
  Environmental Science & Technology, 42(3), 822–830.
- USGS SPARROW resources: <https://www.usgs.gov/mission-areas/water-resources/science/sparrow-modeling>

## Citation

Gorman Sanisaca, Lillian, Hurley, Kyle P., and Alexander, Richard B. (2025).
RSPARROW 2.1.0: An R system for SPARROW modeling. U.S. Geological Survey.

## Disclaimer

This software has been approved for release by the U.S. Geological Survey (USGS).
Although the software has been subjected to rigorous review, the USGS reserves the
right to update the software as needed pursuant to further analysis and review. No
warranty, expressed or implied, is made by the USGS or the U.S. Government as to
the functionality of the software and related material nor shall the fact of release
constitute any such warranty.

Any use of trade, firm, or product names is for descriptive purposes only and does
not imply endorsement by the U.S. Government.
